---
title: "Querying a SOMA experiment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Querying a SOMA experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

In this notebook, we'll take a quick look at how to query a `SOMAExperiment` using the `SOMAExperimentAxisQuery` class. This allows for easy selection of data from a `SOMAMeasurement` by filtering on annotations stored in each axis data frame (i.e., `obs` and `var`).

```{r}
library(tiledbsoma)
```

## Example data

Load the bundled `SOMAExperiment` containing a subsetted version of the 10X genomics [PBMC dataset](https://mojaveazure.github.io/seurat-object/reference/pbmc_small.html) provided by SeuratObject. This will return a `SOMAExperiment` object.

```{r}
experiment <- load_dataset("soma-exp-pbmc-small")
experiment
```

## Querying basics

To perform a query we'll need to initialize a new `SOMAExperimentAxisQuery` object, specifying the `SOMAExperiment` and the `SOMAMeasurement` within the experiment we want to query.

We can see that our current experiment contains only a single measurement: `"RNA"`.

```{r}
experiment$ms
```

Now we can construct our query object.

```{r}
query <- SOMAExperimentAxisQuery$new(
  experiment = experiment,
  measurement_name = "RNA"
)
```

Once it's created, we can use the `query` object to inspect, select, and extract filtered data from the experiment.

For example, we can use `n_obs` and `n_vars` to determine the number of observations and variables that passed our filtering criteria. Since we didn't specify any filtering criteria, these numbers will match the full size of the experiment.

Number of observations:

```{r}
query$n_obs
```

Number of variables:

```{r}
query$n_vars
```

We can also extract any data component from the experiment. Here we'll read in the `obs` data frame from the query.

```{r}
query$obs()
```

## Adding filters

Adding filters requires creating a `SOMAAxisQuery` object that allows you to define coordinates, value filters, or both for an axis.

Here we'll create a query for `obs` that slices the first 40 rows, and then filters that subset based on the `nCount_RNA` column.

```{r}
obs_query <- SOMAAxisQuery$new(
  coords = list(soma_joinid = 0:39),
  value_filter = "nCount_RNA > 100"
)
```

To apply this filter we'll pass it to a new `SOMAExperimentAxisQuery` object.

```{r}
query <- SOMAExperimentAxisQuery$new(
  experiment = experiment,
  measurement_name = "RNA",
  obs_query = obs_query
)
```

Let's see how many observations this query identified.

```{r}
query$n_obs
```

As before, we can load the `obs` data frame into memory but now it only includes the filtered observations.

```{r}
query$obs(column_names = c("obs_id", "nCount_RNA"))$to_data_frame()
```

We can also access the expression data for these filtered cells via `X()`:

```{r}
query$X(layer_name = "counts")
```

## Seurat support

The `query` object also contains methods for loading in the results as a Seurat object (or any of Seurat's component classes).

```{r eval=requireNamespace("SeuratObject", quietly = TRUE)}
query$to_seurat(
  X_layers = c(counts = "counts", data = "data"),
  obs_index = "obs_id",
  var_index = "var_id"
)
```
