---
title: "Selective Updates"
description: >
  How to selectivley update components of an SCGroup.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Updates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

```{r setup}
library(tiledbsc)
library(SeuratObject)
```

# Overview

While *tiledbsc* is currently focused on facilitating the storage and retrieval of complete Seurat/Bioconductor datasets, it is possible to selectively update specific components of an existing `SCGroup`.

# Create the initial dataset

As in the Introduction we'll use the `pbmc_small` dataset from the *SeuratObject* package.

```{r}
data("pbmc_small", package = "SeuratObject")
pbmc_small
```


We'll start by creating a Seurat `Assay` object that contains only raw RNA counts

```{r}
pbmc_small_rna <- CreateAssayObject(
  counts = GetAssayData(pbmc_small[["RNA"]], "counts")
)

Key(pbmc_small_rna) <- Key(pbmc_small[["RNA"]])
```

and then ingest this into a new `SOMA` on disk:

```{r, include = FALSE}
soma <- SOMA$new(uri = file.path(tempdir(), "pbmc_small_rna"))
soma$from_seurat_assay(pbmc_small_rna)
soma
```

This performed the following operations:

- created groups `X`, `obsm`, `varm`, `obsp`, `varp`, and `misc`
- created array, `obs`, with a single dimension containing all cell names and 0 attributes (because Seurat `Assay` objects do not contain cell-level metadata)
- created array, `var`, with a single dimension containing all feature names and 0 attributes (because the feature-level metadata was empty)
- created array `counts` within the `X` group and ingested all data from `pbmc_small_rna`'s `counts` slot

Printing out the `X` group, we can see it contains only a single member, `counts`.

```{r}
soma$X
```

*Note: A Seurat `Assay` instantiated with data passed to the `counts` argument will store a reference to the same matrix in both the `counts` and `data` slots (i.e., `GetAssayData(pbmc_small_rna, "counts")` and `GetAssayData(pbmc_small_rna, "data")`). To avoid redundantly storing data on disk, *tiledbsc* only ingests data from `data` when it is not identical to `counts`.*

# Add new layers for normalized and scaled data

Now let's populate `pbmc_small_rna`'s `counts` slot with normalized counts.

```{r}
pbmc_small_rna <- SetAssayData(
  pbmc_small_rna,
  slot = "data",
  new.data = GetAssayData(pbmc_small[["RNA"]], "data")
)

pbmc_small_rna
```

At this point, `counts` and `data` contain different transformations of the same matrix but only the `counts` data has been written to disk.

We could simply rerun `soma$from_seurat_assay(pbmc_small_rna)` to update the `SCGroup` and ingest the normalized counts matrix into `X/data`, however, this
would perform another full conversion of the entire `Assay` object. As a result, data from the `counts` slot would be *re-ingested* into `X/counts` and feature-level metadata would be re-ingested into `var`.

To avoid these redudndant ingestions, use the `layers` argument to specify which of the assay slots (`counts`, `data`, or `scale.data`) should be ingested. We also set `var = FALSE` to prevent re-ingestion of the feature-level metadata since it hasn't changed.

```{r}
soma$from_seurat_assay(pbmc_small_rna, layers = "data", var = FALSE)
```

Printing out the `X` group again reveals that it now contains 2 members: `counts` and `data`.

```{r}
soma$X
```

Let's repeat this process for the scaled/centered version of the data that's stored in the `scale.data` slot.

```{r}
pbmc_small_rna <- SetAssayData(
  pbmc_small_rna,
  slot = "scale.data",
  new.data = GetAssayData(pbmc_small[["RNA"]], "scale.data")
)

soma$from_seurat_assay(pbmc_small_rna, layers = "scale.data", var = FALSE)
soma$X
```

The `X` group now contains 3 arrays, each of which has only been written to a single time. We can verify this by counting the number of fragments contained within any of individual arrays:

```{r}
soma$X$members$count$fragment_count()
```

TileDB creates a new fragment each time a write is performed, so 1 fragment confirms the array has been written to only once.

# Update an existing layer

If the data in your Seurat `Assay` _has_ changed since it was ingested into TileDB, you can follow the same basic process to update the array.

Here, we'll aritifically shift all of the values in the `scale.data` slot by 1 and then re-ingest the updated data into the `scale.data` layer on disk:

```{r}
pbmc_small_rna <- SetAssayData(
  pbmc_small_rna,
  slot = "scale.data",
  new.data = GetAssayData(pbmc_small[["RNA"]], "scale.data") + 1
)

soma$from_seurat_assay(pbmc_small_rna, layers = "scale.data", var = FALSE)
```

Now let's verify that number of fragments in each `X`'s layers:

```{r}
sapply(
  names(soma$X$members),
  function(x) soma$X$members[[x]]$fragment_count()
)
```


# Session

```{r}
sessionInfo()
```

```{r cleanup, include=FALSE}
tiledb::tiledb_vfs_remove_dir(soma$uri)
```
